---
title: "cu25_deseq2"
output: html_document
date: "2025-07-17"
params:
  quant_dir: "out/salmon"           # directory containing per-sample Salmon dirs with quant.sf
  gtf_file: "data/references/gtf/Mus_musculus.GRCm39.109.gtf"  # path to GTF (default now under data/references)
  project_name: "CU25"              # project prefix for all output filenames
  tximport_rds: !r NULL              # optional precomputed tximport object (.rds)
  sample_table: "CU25_sample_info.csv"  # sample table CSV (rows = samples)
  group_col: "trx"                  # column in sample_table used for design
  out_dir: "out/deseq2"             # output directory for figures/tables
  padj_thresh: 0.05                  # adjusted p-value threshold for summaries/volcano
  lfc_thresh: 0.5                    # absolute log2FC threshold for volcano highlighting
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Preparing Data for DESeq2 with tximport and labeling Gene IDs
```{r}
# -----------------------------
# 1. Load libraries
# -----------------------------
install.packages("BiocManager", repos = "https://cloud.r-project.org")
BiocManager::install("tximport")
BiocManager::install("GenomicFeatures")
BiocManager::install("txdbmaker")

library(tximport)
library(readr)
library(GenomicFeatures)
library(txdbmaker)
library(dplyr)
library(tibble)

# Parameters (backward-compatible defaults in YAML)
quant_dir    <- params$quant_dir
gtf_file     <- params$gtf_file
out_dir      <- params$out_dir
padj_thresh  <- params$padj_thresh
lfc_thresh   <- params$lfc_thresh
tximport_rds <- params$tximport_rds
project_name <- params$project_name
file_prefix  <- paste0(project_name, "_")

dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# -----------------------------
# 2. Define file paths 
# -----------------------------
# Update this to the actual path of your GTF and quant directories
# (now parameterized via YAML params above)
# gtf_file <- "reference/Mus_musculus.GRCm39.109.gtf"  # kept default via params
quant_dirs <- list.dirs(quant_dir, recursive = FALSE, full.names = TRUE)

# -----------------------------
# 3. Create transcript → gene map from GTF
# -----------------------------
txdb <- makeTxDbFromGFF(gtf_file, format = "gtf")
k <- keys(txdb, keytype = "TXNAME")
tx2gene <- AnnotationDbi :: select(txdb, 
                                   keys = k, 
                                   columns = c("GENEID"), 
                                   keytype = "TXNAME")

# Optional: convert to clean tibble
tx2gene <- as_tibble(tx2gene)

# -----------------------------
# 4. Read Salmon quant.sf files (or load provided tximport RDS)
# -----------------------------
if (!is.null(tximport_rds)) {
  txi <- readRDS(tximport_rds)
} else {
  # Each folder in quant_dirs should contain a quant.sf file
  files <- file.path(quant_dirs, "quant.sf")
  names(files) <- basename(quant_dirs)

  # Import transcript-level Salmon quantification, summarize to gene-level
  txi <- tximport(files, 
                  type = "salmon", 
                  tx2gene = tx2gene,
                  ignoreTxVersion = TRUE)
}

# -----------------------------
# 5. Add Gene Symbols (Optional)
# -----------------------------
# Use biomaRt to annotate gene IDs with symbols (requires internet)
if (!requireNamespace("biomaRt", quietly = TRUE)) {
  install.packages("biomaRt")
}
library(biomaRt)

ensembl <- useEnsembl(biomart = "genes", dataset = "mmusculus_gene_ensembl")
gene_ids <- rownames(txi$counts)

annot <- getBM(attributes = c("ensembl_gene_id", "external_gene_name"),
               filters = "ensembl_gene_id",
               values = gene_ids,
               mart = ensembl)

# Merge gene symbols
counts_with_symbols <- as.data.frame(txi$counts)
counts_with_symbols$ensembl_gene_id <- rownames(counts_with_symbols)
counts_annotated <- dplyr::left_join(counts_with_symbols, annot, by = "ensembl_gene_id")

# Move gene symbol to front
counts_annotated <- counts_annotated %>%
  dplyr::select(external_gene_name, dplyr::everything(), -ensembl_gene_id) %>%
  dplyr::rename(gene_symbol = external_gene_name)


# -----------------------------
# 6. Save outputs
# -----------------------------

# Rename columns to explicit sample IDs from the sample table
sample_table_path <- params$sample_table
st <- read.csv(sample_table_path, row.names = 1, check.names = FALSE)
sample_ids <- rownames(st)

clean_names_from_sample_table <- function(current_names, sample_ids) {
  vapply(current_names, function(nm) {
    hits <- sample_ids[sapply(sample_ids, function(s) grepl(s, nm, fixed = TRUE))]
    if (length(hits) == 1) hits else nm
  }, character(1))
}

# Apply to txi counts and annotated counts
colnames(txi$counts) <- clean_names_from_sample_table(colnames(txi$counts), sample_ids)

library(dplyr)

counts_annotated <- counts_annotated %>%
  dplyr::select(gene_symbol, dplyr::everything())

count_cols <- setdiff(colnames(counts_annotated), "gene_symbol")
new_count_cols <- clean_names_from_sample_table(count_cols, sample_ids)
colnames(counts_annotated)[match(count_cols, colnames(counts_annotated))] <- new_count_cols

# Reorder to match sample table first, then any remaining
ordered_cols <- c("gene_symbol",
                  sample_ids[sample_ids %in% new_count_cols],
                  setdiff(new_count_cols, sample_ids))
counts_annotated <- counts_annotated[, ordered_cols, drop = FALSE]

write.csv(counts_annotated, file = file.path(out_dir, paste0(file_prefix, "gene_counts_with_symbols.csv")), row.names = FALSE)
write.csv(txi$counts, file = file.path(out_dir, paste0(file_prefix, "raw_gene_counts.csv")))

cat("✔ Gene-level counts saved as:\n- ", file.path(out_dir, paste0(file_prefix, "gene_counts_with_symbols.csv")), "\n- ", file.path(out_dir, paste0(file_prefix, "raw_gene_counts.csv")), "\n", sep = "")
```


# Loading packages

```{r}

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
install.packages("BiocManager", repos = "https://cloud.r-project.org")
BiocManager::install("DESeq2")

install.packages("tidyverse")

library(tidyverse)
library(DESeq2)
library(dplyr)

```

# Downloading dataset and cleaning

```{r}
# project name prefix for outputs
name <- project_name

# Read count matrix and sample info
df <- read.csv(file.path(out_dir, paste0(file_prefix, "gene_counts_with_symbols.csv")), stringsAsFactors=FALSE)

# removing blank rows
df <- df[df[[1]] != "",]


# salmon spits out estimates, which are non-integer counts. DESeq2 requires integer counts. look into alternatives for differential expression. can also load the txi object directly into DESeqDataSetFromTximport
df <- df %>%
  mutate(across(where(is.numeric), ~ round(.)))

# sample data import
colData <- read.csv(params$sample_table, row.names = 1)

# separating out the C26 dataset
df <- df %>%
  dplyr::select(c("gene_symbol", rownames(colData)))

# removing NA values
df <- df[!is.na(df$gene_symbol), ]

# creating counts matrix with genes as rowname
counts_data <- df[,-1]
rownames(counts_data) <- make.unique(as.character(df[[1]]))

# making sure the row names in colData matches to column names in counts_data
all(colnames(counts_data) %in% rownames(colData))
all(colnames(counts_data) == rownames(colData))
```


# Contructing a DESeqDataSet object

```{r}
design_formula <- as.formula(paste("~", params$group_col))
dds<- DESeqDataSetFromMatrix(countData = counts_data,
                       colData = colData,
                       design = design_formula)
dds

# pre-filtering: removing rows with low gene counts
# keeping rows that have at least 10 reads total
dds <- dds[ rowSums(counts(dds)) >= 10, ]

# set the factor level
dds[[params$group_col]] <- relevel(dds[[params$group_col]], ref = "control")
```

#NOTE: if you have technical replicates collapse them
```

# Run DESeq

```{r}
dds <- DESeq(dds)

res <- results(dds)
res

# baseMean = average of normalized counts
# stat = Waldtest stats

# saving DEGs to data table for further analysis and saving data_table to csv
res_df <- as.data.frame(res)

write.csv(res_df,
          file = file.path(out_dir, paste0(file_prefix, "DESeq2_full_results.csv")),
          row.names = TRUE)
```

# Explore results

```{r}
summary(res)
# upregulated/downregulated genes, outliers. Default padj is .1

# increasing cutoff to < params$padj_thresh
res0.05 <- results(dds, alpha = padj_thresh)
summary(res0.05)

# contrasts
resultsNames(dds)

# MA plot
plotMA(dds)
```

# Volcano plot

```{r}
# Make sure you have ggrepel installed
if (!requireNamespace("ggrepel", quietly=TRUE)) {
  install.packages("ggrepel")
}
library(ggplot2)
library(ggrepel)

# Prepare dataframe
res_df$gene <- rownames(res_df)

# Add column to specify up vs downregulated
res_df$diffexpressed <- "NO"
res_df$diffexpressed[res_df$log2FoldChange >  lfc_thresh & res_df$padj < padj_thresh] <- "Up"
res_df$diffexpressed[res_df$log2FoldChange < -lfc_thresh & res_df$padj < padj_thresh] <- "Down"

# Gene labels (top 30)
res_df$delabel <- ifelse(res_df$gene %in% head(res_df[order(res_df$padj), "gene"], 30), res_df$gene, NA)

volcano_plot <- ggplot(data=res_df, aes(x= log2FoldChange, y=-log10(padj), col = diffexpressed, label = delabel)) +
  geom_vline(xintercept = c(-lfc_thresh, lfc_thresh), col = "gray", linetype = "dashed")+
  geom_hline(yintercept = -log10(padj_thresh), col = "gray", linetype = "dashed")+
  geom_point(size =.5)+
  scale_color_manual(values = c("#00AFBB", "grey", "red"),
                     labels = c("Downregulated", "Not Significant", "Upregulated")) +
  labs(color = 'Up vs Down', x = expression("log"[2]*"FC"), y = expression("-log"[10]*"padj")) +
  ggtitle("Control vs ID8 (p53-/-)") +
  geom_text_repel(max.overlaps = Inf)

ggsave(file.path(out_dir, paste0(file_prefix, "volcano_plot.pdf")), 
       plot = volcano_plot,
       device = pdf,
       width = 10,
       height = 8)
```

# VST normalized count files and PCA plot

```{r}
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = params$group_col)

write.csv(assay(vsd), file = file.path(out_dir, paste0(file_prefix, "vst_norm_counts.csv")))

pca_plot <- plotPCA(vsd, intgroup = params$group_col)+ stat_ellipse(linetype = 3)

ggsave(file.path(out_dir, paste0(file_prefix, "PCA_plot.pdf")), 
       plot = pca_plot,
       device = pdf,
       width = 7,
       height = 5)
```

# Enricher

```{r}

# libraries
library(enrichR)
library(writexl)
library(openxlsx)

# Defining Databases
db <- c("KEGG_2019_Mouse", "Reactome_2022", "GO_Biological_Process_2025", "GO_Cellular_Component_2025", "GO_Molecular_Function_2025", "Tabula_Muris", "Elsevier_Pathway_Collection", "Bioplanet_2019")

# Subsetting DEGs to only upregulated using only top 200 genes
goi <- res_df %>%
  dplyr::filter(padj < padj_thresh, log2FoldChange > 1.5) %>%
  dplyr::slice_min(order_by = padj, n = 200)

# Extracting gene list as character vector
genes <- as.character(goi$gene)

# Enrichment
edb<- enrichr(genes, db)


# Creating Excel workbook
wb <-openxlsx::createWorkbook()

for (db_name in db){
  addWorksheet(wb, sheetName = db_name)
  writeData(wb,
            sheet = db_name,
            x = edb[[db_name]],
            rowNames = FALSE)
}

# saving workbook
saveWorkbook(wb,
             file = file.path(out_dir, paste0(file_prefix, "Upregulated_cachexia_Enrichment_results.xlsx")),
             overwrite = TRUE)


# Subsetting DEGs to only downregulated using only top 200 genes
goi <- res_df %>%
  dplyr::filter(padj < padj_thresh, log2FoldChange < -1.5) %>%
  dplyr::slice_min(order_by = padj, n = 200)

# Extracting gene list as character vector
genes <- as.character(goi$gene)

# Enrichment
edb<- enrichr(genes, db)


# Creating Excel workbook
wb <-openxlsx::createWorkbook()

for (db_name in db){
  addWorksheet(wb, sheetName = db_name)
  writeData(wb,
            sheet = db_name,
            x = edb[[db_name]],
            rowNames = FALSE)
}

# saving workbook
saveWorkbook(wb,
             file = file.path(out_dir, paste0(file_prefix, "Downregulated_cachexia_Enrichment_results.xlsx")),
             overwrite = TRUE)

```
